---
title: "Using Machine Learning Estimators to Predict Age of Death"
authors: Nathan Seltzer (UC Berkeley) and Casey Breen (UC Berkeley)
date: October 10, 2022
---

## Description
 
Can age of death be predicted using Census Data? We test this question using linked historical administrative data from the CenSoc project, which combines the full count 1940 Census from IPUMS USA with Social Security Death Records that span the years 1988-2005. We leverage several ML algorithms and approximately 150 features to generate predictions of age of death for birth cohorts born at the beginning of the 20th century. The findings from this research emphasize the promises and perils of using statistical and ML models to attempt to answer one of the most fundamental questions: can death be predicted?

## Workflow:

(a) Import Data
(b) Clean Data
(c) Feature Engineering
(d) ML Model Comparison
(e) Bootstrap Best Model


```{r}
## library packages 
library(tidyverse) # for various helper functions, pipes, etc.
library(data.table) # for quick import/export of data
library(caret) # for ML model auto-tuning
library(doParallel) # for parallel processing
library(ipumsr) # for variable labels
library(tidymodels) # for data partition
library(here)

`%notin%` <- Negate(`%in%`)

### Import 
numident_linked <- fread("/data/josh/CenSoc/censoc_data/censoc_linked_to_census/v2.1/censoc_numident_v2.1_linked.csv")

## select one cohort 
numident_linked_1910 <- numident_linked %>%
  filter(link_abe_exact_conservative == 1) %>%
  filter(byear == 1910) # %>%
# filter(race_first_cyear < 1988, race_last_cyear < 1988) # %>%
#sample_frac(.2)


## select seeds 
set.seed(2232)

## split data 
split <- initial_split(numident_linked_1910, prop = 0.75) # set to .6
train_data <- training(split)
test_data <- testing(split)
```


## Functions 

```{r}
# Recode function - converts numeric to factor, sets missing values to NA
recodes <- function(df) {
  df2 <- df %>%
    mutate(
      # SEX = factor(SEX),
      # MARST = factor(ifelse(MARST == 3, 4, MARST)), # set separated to divorced
      # RACE = factor(RACE),
      # METRO = factor(METRO),
      # HISPAN = factor(HISPAN),
      # CLASSWKR = factor(CLASSWKR),
      # LABFORCE = factor(LABFORCE),
      # URBAN = factor(URBAN),
      # FARM = factor(FARM),
      # OWNERSHP = factor(OWNERSHP),
      EDUCD = ifelse(EDUCD == 999, NA, EDUCD),
      INCWAGE = ifelse(INCWAGE == 999998 | INCWAGE == 999999, NA, INCWAGE),
      RENT = ifelse(RENT == 9998 | RENT == 9999, NA, RENT))
  
  return(df2)
}

# missing model creator function - models missing data using lm
missing_model_creator <- function(train, test){
  
  registerDoSEQ()
  unregister <- function() {
    env <- foreach:::.foreachGlobals
    rm(list=ls(name=env), pos=env)
  }
  
  m_EDUCD <- train(form = EDUCD ~ SEX + INCWAGE + SEI, data = train, method = "lm", na.action = na.omit)
  m_incwage <- train(form = INCWAGE ~ SEX + EDUCD + SEI, data = train, method = "lm", na.action = na.omit)
  m_rent <- train(form = RENT ~ SEX + EDUCD + INCWAGE + SEI, data = train, method = "lm", na.action = na.omit)

  train$EDUCD = ifelse(is.na(train$EDUCD)==TRUE, round(predict(m_EDUCD, train), 0), train$EDUCD)
  train$INCWAGE = ifelse(is.na(train$INCWAGE)==TRUE, round(predict(m_incwage, train), 0), train$INCWAGE)
  train$RENT = ifelse(is.na(train$RENT)==TRUE, round(predict(m_rent, train), 0), train$RENT)

  test$EDUCD = ifelse(is.na(test$EDUCD)==TRUE, round(predict(m_EDUCD, test), 0), test$EDUCD)
  test$INCWAGE = ifelse(is.na(test$INCWAGE)==TRUE, round(predict(m_incwage, test), 0), test$INCWAGE)
  test$RENT = ifelse(is.na(test$RENT)==TRUE, round(predict(m_rent, test), 0), test$RENT)

  train_test <- list(train,test)
  
  return(train_test)
}
```



```{r}
# X is our training sample.
x_train = train_data %>%
  select(EDUCD, SEI, INCWAGE, OCCSCORE, PERNUM, NUMPERHH, SEX, RACE, METRO,
         PRESGL, MARST, CITYPOP, NSIBS, RENT, HISPAN, CLASSWKR, MOMLOC, MIGRATE5,
         LABFORCE, NFAMS, URBAN, FARM, OWNERSHP, PRESGL, NCOUPLES, NMOTHERS,
         NFATHERS, MULTGEND, WKSWORK1, WKSWORK2, HRSWORK1, HRSWORK2, MARRNO,
         AGEMARR, CHBORN, EMPSTAT) 

# Create a holdout set for evaluating model performance.
# Note: cross-validation is even better than a single holdout sample.
x_holdout = test_data %>%
  select(EDUCD, SEI, INCWAGE, OCCSCORE, PERNUM, NUMPERHH, SEX, RACE, METRO,
         PRESGL, MARST, CITYPOP, NSIBS, RENT, HISPAN, CLASSWKR, MOMLOC, MIGRATE5,
         LABFORCE, NFAMS, URBAN, FARM, OWNERSHP, PRESGL, NCOUPLES, NMOTHERS,
         NFATHERS, MULTGEND, WKSWORK1, WKSWORK2, HRSWORK1, HRSWORK2, MARRNO,
         AGEMARR, CHBORN, EMPSTAT)  


x_train2 <- recodes(x_train)
x_holdout2 <-recodes(x_holdout)
train_test <- missing_model_creator(x_train2, x_holdout2)

normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

x_train3 <- data.frame(train_test[1]) %>%
  mutate_if(is.numeric, normalize)
x_holdout3 <- data.frame(train_test[2]) %>%
  mutate_if(is.numeric, normalize)

y_train = train_data %>% select(death_age)
y_train = as.vector(y_train$death_age)

y_holdout = test_data %>% select(death_age)
y_holdout = as.vector(y_holdout$death_age)

## save data 
write_rds(x = y_train, file = here("data/y_train.rds"))
write_rds(x = y_holdout, file = here("data/y_holdout.rds"))

write_rds(x = x_train3, file = here("data/x_train.rds"))
write_rds(x = x_holdout3, file = here("data/x_holdout.rds"))

```




